---
title: "Preliminary Data Exploration"
output:
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

This R Markdown document details the steps of preliminary exploratory analysis conducted on the MAARS dataset for preparing for the analysis to determine the connections between atopic dermititis lesioned and non-lesioned skin samples. 


## Preparation

Load libraries, load data, and prepare the data. 

```{r prep, message=FALSE}
# load libraries
#library(rstudioapi)
library(ggplot2)
library(dplyr)
library(qdapTools)
library(kableExtra)

# set working directory
#setwd(dirname(rstudioapi::getSourceEditorContext()$path))
setwd("/Users/thewooz/Documents/Omics-Project")

# load the data
all_data<- read.csv("/Users/thewooz/Documents/Omics-Project/filter-data/all.csv")
ad_full<-  read.csv("/Users/thewooz/Documents/Omics-Project/filter-data/ad_full.csv")

# order the clinical groups
all_data$clinical_group<- factor(all_data$clinical_group, levels=c("AD","PSO","CTRL"))

# rename the factors for lesional/non-lesional
levels(all_data$lesional) <- c("Lesional","Non-Lesional")

# reduce the all_data file to include just the AD samples
ad_data<- subset(all_data, all_data$clinical_group == 'AD')
```


## Describe the Samples

Counts of lesional vs. non-lesional samples among each of the clinical groups.

```{r describe, message=FALSE}
# table of percent of SAMPLES per clinical group and lesional/non-lesional
clingroup<- round(prop.table(table(all_data$clinical_group, all_data$lesional))*100,2)
clingroup %>% kable(caption = "Percent of Lesional Samples per Clinical Group") %>% kable_styling(bootstrap_options = "striped", full_width = F)

# bar chart for SAMPLES per clinical group and lesional/non-lesionsl
ggplot2::ggplot(data=all_data, aes(clinical_group, fill=lesional))+
  geom_bar(stat="count", position="stack")+
  scale_fill_manual(values=c("red4","steelblue4"),name="Lesional/Non-Lesional", labels=c("Lesional","Non-Lesional"))+
  theme_minimal()+
  xlab("Clinical Group")+
  ylab("Number of Samples")+
  ggtitle("Number of Lesional Samples per Clinical Group")
```


## Describe the Participants

Create a dataset of unique participants in all clinical groups and describe their characteristics.

```{r participants, message=FALSE}
# create dataset of unique participants
reduce<- all_data[ ,c(6,2,7:34)]
undup_pts<- reduce[!duplicated(reduce$MAARS_identifier), ]

# report up to 4 digits
options(digits=4)

# gender (percent)
gender<- 
  undup_pts %>%
  group_by(clinical_group, Gender) %>%
  summarise(n = n()) %>% 
  mutate(RelativeFrequency = paste0(round(100 * n/sum(n), 2), "%"))
gender %>% kable(caption = "Gender") %>% kable_styling(bootstrap_options = "striped", full_width = F)

# age (min, max, mean, SD)
age<-
  undup_pts %>%
  group_by(clinical_group) %>%
  summarize(Min=min(CUSTOM_Age, na.rm=TRUE),
            Max=max(CUSTOM_Age, na.rm=TRUE), 
            Mean=mean(CUSTOM_Age, na.rm=TRUE), 
            SD=sd(CUSTOM_Age, na.rm=TRUE))
age %>% kable(caption = "Age") %>% kable_styling(bootstrap_options = "striped", full_width = F)

# boxplot for age
ggplot2::ggplot(data=undup_pts, aes(x=clinical_group, y=CUSTOM_Age))+
  geom_boxplot()+
  theme_minimal()+
  xlab("Clinical Group")+
  ylab("Age")+
  ggtitle("Age per Clinical Group")

```

## Race of AD Group Patients

```{r race}
race_tab<- as.data.frame(table(ad_full$patient.Diagnostic...Phenotypic.Data.Ethnicity.Family.History.Ethnicity..Ethnicity.))
options(digits = 3)
race_tab<- mutate(race_tab, Percent=Freq/sum(Freq)*100)
race_tab %>% kable(caption = "Race (AD Group)") %>% kable_styling(bootstrap_options = "striped", full_width = F)


```



##  Clinical Characteristics of the AD Group

```{r clinical, message=FALSE}
# create dataframe of unique AD patients
undup_AD<- undup_pts[which(undup_pts$clinical_group=='AD'), ]

# number/percent of AD patients with each type of allergy
allergies<- (mtabulate(undup_AD[grep('Allergies',names(undup_AD), value=T)]))
allergies$Percent<- allergies[,2]/allergies[,1]+allergies[,2]
row.names(allergies)<- c("Psuedo Drug","Dust Mite","Food","Pollen","Contact", "Drug","Animal")
p_allergies<-allergies[ ,2:3]
colnames(p_allergies)[colnames(p_allergies)=="True"]<- "n"
p_allergies %>% kable(caption = "Allergies (AD Group)") %>% kable_styling(bootstrap_options = "striped", full_width = F)

# number/percent of AD patients taking each type of medication
meds<- (mtabulate(undup_AD[grep('Medication',names(undup_AD), value=T)]))
meds$percent<- meds[,2]/meds[,1]+meds[,2]
row.names(meds)<- c("Anti-Hypertensive","Anti-Inflammatory","Other Hormones","Thyroid Hormones","Statins", "Insulin","Other")
p_meds<- meds[ ,2:3]
colnames(p_meds)[colnames(p_meds)=="True"]<- "n"
p_meds %>% kable(caption = "Medications (AD Group)") %>% kable_styling(bootstrap_options = "striped", full_width = F)

# number/percent of AD patients with each type of chronic disease
chrondis<- (mtabulate(undup_AD[grep('chronic',names(undup_AD), value=T)]))
chrondis$percent<- chrondis[,2]/chrondis[,1]+chrondis[,2]
row.names(chrondis)<- c("Hyperlipidemia","Other","Diabetes (non-insulin)","Thyroid Dysfuntion","Asthma", "Hypertension")
p_chrondis<- chrondis[ ,2:3]
colnames(p_chrondis)[colnames(p_chrondis)=="True"]<- "n"
p_chrondis %>% kable(caption = "Chronic Disease (AD Group)") %>% kable_styling(bootstrap_options = "striped", full_width = F)

# number/percent of AD patients with skin malignancy or other malignancy
malig<- (mtabulate(undup_AD[grep('Malig',names(undup_AD), value=T)]))
malig$percent<- malig[,2]/malig[,1]+malig[,2]
row.names(malig)<- c("Skin malignancy","Other malignancy")
p_malig<- malig[ ,2:3]
colnames(p_malig)[colnames(p_malig)=="True"]<- "n"
p_malig %>% kable(caption = "Malignancies (AD Group)") %>% kable_styling(bootstrap_options = "striped", full_width = F)

# number/percent of AD patients with a family history of AD
famhist<-as.data.frame(addmargins(table(undup_AD$CUSTOM_Fam._hist._Atopic_dermatitis)))
famhist$Percent<- (famhist$Freq/famhist[4,2])*100
famhist<- famhist[2:3, ]
famhist$Var1<- as.factor(c("Family history of AD", "No family history of AD"))
colnames(famhist)[colnames(famhist)=="Freq"]<- "n"
colnames(famhist)[colnames(famhist)=="Var1"]<- ""
rownames(famhist) <- NULL
famhist %>% kable(caption = "Family History (AD Group)") %>% kable_styling(bootstrap_options = "striped", full_width = F)
```



## Interpretation

Patients in the AD group are more likely to be male (55.8%). The psoriasis group is also predominately male (78.5%) while the control group is predominantly female (63.7%).

The mean age of AD participants is nearly 44 years, with a minimum age of 20 and a maximum age of 83. The psoriasis group is slightly older on average (mean=49) and the control group is generally younger (mean=35).

A substantial portion of the AD group report allergies to substances, including pollen (48.2%), food (37.8%), and animals (33.6%). This information may be of imporantance since people with AD are more likely to develop hay fever and are advised to avoid allergens in an effort to prevent a flare-up of eczema. ^[McIntosh,J. (2017, Nov 14). What's to know about eczema? https://www.medicalnewstoday.com/articles/14417.php.]

Just under half (43.0%) of the AD group report a family history of AD. 
  




