---
title: "Quality Check"
output:
  html_document:
    theme: united
    toc: yes
    toc_float:
      collapsed: no
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Introduction

Although the data provided was previously normalized, it will be checked for normality, a dendogram will be created, and PCA will be conducted. 

## Preparation

Load libraries, load data, and prepare the data. 

```{r prep, message=FALSE}
# load libraries
#library(limma)
#library(data.table)
#library(gdata)

# set working directory
setwd("/Users/thewooz/Documents/Omics-Project")

# read in sample information, keeping those that are AD
samples <- read.csv('/Users/thewooz/Documents/Omics-Project/filter-data/all.csv')
samples_AD <- subset(samples, samples$clinical_group == 'AD')

# read in transcriptome data, change row names and remove the var 'X'
transcriptome <- read.csv('/Users/thewooz/Documents/Omics-Project/filter-data/transcriptome.csv')
rownames(transcriptome) <- transcriptome$X
transcriptome <- transcriptome[,-c(1)]

# keep the transcriptomes that are AD
transcriptome_AD <- transcriptome[rownames(transcriptome) %in% samples_AD$sample_id, ]

# create an index, matching the sample_id to the row in the transcriptome
indices <- match(samples_AD$sample_id, rownames(transcriptome_AD))
# stop the process if there's any lines that have NA (sample_id has not matching transcriptome)
stopifnot(sum(is.na(indices)) == 0)

# order the transcriptome data according to the vector
transcriptome_AD <- transcriptome_AD[indices,]
# stop if not every row matches a sample_id
stopifnot(all(rownames(transcriptome_AD) == samples_AD$sample_id))
# transpose transcriptome data
transcriptome_AD <- t(transcriptome_AD)

# recode lesional/non-lesional
samples_AD$lesional_new[samples_AD$lesional=="LES"]<- 1
samples_AD$lesional_new[samples_AD$lesional=="NON_LES"]<- 0   
```


## Verify Normality of the Samples

```{r normality}
bp_col<- samples_AD$lesional
levels(bp_col) <- c("red4", "steelblue4")
boxplot(transcriptome_AD[,-1], col=as.character(bp_col), 
        xlab="Sample", ylab="Intensity", axes=F)
axis(2)
box()
```


## Dendrogram of AD Genes

Create a dendrogram of the genes sampled for the AD group. 


```{r}
dist_data<- dist(t(transcriptome_AD))
plot(flashClust::hclust(dist_data, method="ward"), labels=samples_AD$lesional)

```


## Principle Component Analysis (PCA)



## Heatmap Clustering


